---
title: "HW3-am6592"
author: "Adeena Moghni"
date: "2024-10-07"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse) #read csv
library(readxl) #read excel
library(haven) #read SAS
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
  fig.align = "center"
)
```

## Problem 2

```{r}
demo_df = 
  read_csv(file = "data_imports/nhanes_covar.csv", 
          na = c(".", "NA", ""),
          skip = 4,
          col_types = cols(
    `SEQN` = col_character(),
    `sex` = col_factor(),
    `age` = col_integer(),
    `BMI` = col_double(),
    `education` = col_factor())) %>% 
  janitor::clean_names()

mims_df = 
  read_csv(file = "data_imports/nhanes_accel.csv", 
           na = c(".", "NA", ""),
           col_types = cols (
             `SEQN` = col_character() )) %>% 
  janitor::clean_names()

full_accel_df = left_join(mims_df, demo_df, by = "seqn") %>% 
  filter(age > 21, !is.na(seqn), !is.na(sex), !is.na(age), !is.na(bmi), !is.na(education)) %>% 
  pivot_longer(
    min1:min1440,
    names_to = "minute", 
    values_to = "MIMS",
    names_prefix = "min") %>%  
  mutate(
    minute = as.factor(minute)
  )
```

```{r}
education_count =
  full_accel_df %>% 
  distinct(seqn, .keep_all = TRUE) %>% 
  mutate(sex = ifelse(sex == 1, "Male", "Female")) %>% 
  mutate(education = ifelse(education == 1, "Less than High School",
                            ifelse(education == 2, "High School Equivalent", "More than High School"))) %>% 
  group_by(sex, education)%>% 
  summarize(counts = n())%>% 
  pivot_wider(
    names_from = "education", 
    values_from = "counts" ) %>% 
  select(1, 3, 2, 4)

knitr::kable(education_count)
```

```{r}
demo_plot = 
  demo_df %>% 
  filter(age > 21) %>%
  na.omit(demo_df) %>% 
  mutate(sex = ifelse(sex == 1, "Male", "Female")) %>% 
  mutate(education = ifelse(education == 1, " Less than High School",
                            ifelse(education == 2, "High School Equivalent", "More than High School")))

ggplot(demo_plot, aes(x = age, fill = education)) +
  geom_density(alpha = .4, adjust = .5, color = "blue") +
  facet_grid(~sex) +
  labs(title = "Distribution of Education Levels by Age",
       x = "Age (years)", y = "density",
       fill = "Education Level") +
  theme_bw()
```
The number of males and females who have less than a high school equivalent education are about the same (27 and 28, respectively), as are the number of men and women who have more than a high school equivalent education (54 and 59, respectively). However, the number of males who have a high school equivalent education far outnumber the number of women who have the same education level (34 and 23 , respectively). In both sexes, about half the participants have more than a high school education. When blocking for age, those with more than a high school equivalent education tended to skew towards younger women from the ages fo 20 to 40. Those with a high school equivalent and less than high school valent education tended to skew towards older women (aged 60+) but there was local maxima for both these education levels in middle aged women (40-60 years old), but not as pronounced compared to in older women. In men, those with more than a high school equivalent education also tended to be younger, but this skewness was not as pronounced as compared to the distribution of women of the same education level. Those with less than a high school equivalent education skewed to be either between the ages of 40 to 50 or 60+, while those with a high school equivalent education skewed to middle ages men (between 50 to 60 years old)
```{r}
activity_plot = 
  full_accel_df %>%
  group_by(seqn, age, sex, education) %>% 
  summarize(total_activity = sum(MIMS)) %>% 
  mutate(sex = ifelse(sex == 1, "Male", "Female")) %>% 
  mutate(education = ifelse(education == 1, " Less than High School",
                            ifelse(education == 2, "High School Equivalent", "More than High School")))
         

ggplot(activity_plot, aes(x = age, y = total_activity, color = education)) +
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) +
  facet_grid(~sex) +
  labs(title = "Total Daily Activity Levels vs. Age",
       x = "Age (years)", y = "Total Daily Activity (*MIMS)",
       color = "Education Level") + 
  theme_bw()
```

```{r}
ggplot(activity_plot, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) +
  facet_grid(~education) +
  labs(title = "Total Daily Activity Levels vs. Age",
       x = "Age (years)", y = "Total Daily Activity (*MIMS)",
       color = "Education Level") +
  theme_bw()
```

## Problem 3
```{r message = FALSE}
jan2020_df = 
  read_csv(file = "data_imports/citibike/Jan_2020_Citi.csv", na = c(".", "NA", "")) %>% 
  mutate("date" = "January 2020")

jan2024_df = 
  read_csv(file = "data_imports/citibike/Jan_2024_Citi.csv", na = c(".", "NA", ""))%>% 
  mutate("date" = "January 2024")

july2020_df = 
  read_csv(file = "data_imports/citibike/July_2020_Citi.csv", na = c(".", "NA", ""))%>% 
  mutate("date" = "July 2020")

july2024_df = 
  read_csv(file = "data_imports/citibike/July_2024_Citi.csv", na = c(".", "NA", ""))%>% 
  mutate("date" = "July 2024")

combined_df = 
  bind_rows(jan2020_df, july2020_df, jan2024_df, july2024_df) %>% 
  rename("member_type" = "member_casual")
```

```{r}
ride_count =
  combined_df %>% 
  na.omit(combined_df) %>%
  group_by(date, member_type) %>%
  summarize(counts = n()) %>% 
  pivot_wider(
    names_from = "member_type", 
    values_from = "counts" )

knitr::kable(ride_count)
  
```

```{r}
july2024_rank =
  july2024_df %>% 
  na.omit(july_df) %>%
  group_by(start_station_name) %>%
  summarize(counts = n()) %>% 
  arrange(desc(counts)) %>% 
  slice(1:5)

knitr::kable(july2024_rank)

```

```{r}
median_plot = 
  combined_df %>% 
  na.omit(combined_df) %>%
  separate(date, into = c("month", "year")) %>% 
  group_by(weekdays, month, year) %>% 
  summarize(median_ride = median(duration))
  
ggplot(median_plot, aes(x = median_ride, y = weekdays, fill = year)) +
  geom_histogram(stat = "identity", position = "dodge") +
  facet_grid(~month) +
  labs(title = "Median Ride Time (minutes) vs. Days of the Week",
       x = "Median Ride Time (minutes)", y = "Weekdays") +
  theme_bw()
  
```


```{r}
ride_dist_plot = 
  combined_df %>% 
  na.omit(combined_df) %>%
  separate(date, into = c("month", "year")) %>% 
  filter(year == 2024) %>% 
  group_by(month, member_type, rideable_type) %>% 
  summarize(duration)

ggplot(ride_dist_plot, aes(x = month, y = duration, color = member_type)) +
  geom_boxplot() +
  facet_grid(~rideable_type) + 
  labs(title = "Distribution of Ride Times (minutes) vs. Month",
       x = "Month", y = "Ride Duration (minutes)") +
  theme_bw()
```

